Supplementary Information for:Long- and short-read metabarcoding technologies reveal similar spatio-temporal structures in fungal communities

### 2021-03-16 writing/transect_supplement_rev2.tex
output diff from second revision (to make sure there are no regressions from accepted version)

% full page landscape figures
\usepackage{rotating}
% \DeclareDelayedFloatFlavour{sidewaysfigure}{figure}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[style=apa,]{biblatex}
\addbibresource{all.bib}
\addbibresource{R.bib}

@@

\textbf{Supplementary file 3}: ML tree of all long metabarcoding ASVs (ITS1--5.8S--ITS2--LSU) assigned to kingdom Fungi.
Thick branches represent rapid bootstrap values \textgreater90\%.
Taxonomic assignment by PHYLOTAX, based on assignments using RDPC, SINTAX, and IDTAXA, with Unite, Warcup, and RDP reference databases.
Taxon names in blue were not fully resolved with PHYLOTAX, and are given as a list of alternative assignments, along with the number of methods which gave each assignment (max 9).
Taxon names in red were reconstructed as polyphyletic.
In some cases, this may be due to a poorly identified sequence ``splitting'' the group (e.g., the single \emph{Monascus} sequence nested in \emph{Aspergillus}; Aspergillaceae and Trichocomaceae sequences nested in Onygenales), but in other cases, the tree has clearly failed to reconstruct groups which are generally considered to be monophyletic based on multi-gene trees (e.g., Dothidiomycetes, Cantharellales).

@@

\hypertarget{Sampling-supplement}{%
\subsection{Sampling}\label{Sampling-supplement}}

Sampling was conducted at two sites (Ang: N 9.75456° E 2.14064°; Gan: N 9.75678° E 2.31058°) approximately 30~km apart in the \emph{Forêt Classée de l'Ouémé Supérieur} (Upper Ouémé Forest Reserve) in central Benin.
Both sites were located in West Sudanian savannah woodlands \autocite{olson2001,yorou2014} dominated by the ECM host tree \emph{Isoberlinia doka} (Caesalpinioideae).
At each site, 25~soil samples were collected along a single 24~m linear transect at intervals of 1~m in May 2015.
One third of the sample locations (3~m spacing) were resampled one year later in June 2016, for a total of 67 samples.
For each sample, any coarse organic debris was removed from the soil surface and a sample of approximately 5~cm~× 5~cm~× 5~cm was extracted with a sterilized knife blade.
Each sample was sealed in a plastic zipper bag and homogenized by shaking and manually breaking apart soil aggregations.
Approximately 250~mg total of soil was collected from two locations in the homogenized soil sample and placed into a separate 2.0~mL microtube containing 750~mL of field lysis/preservation buffer and lysis beads (Xpedition\textsuperscript{TM} Soil/Fecal DNA miniprep, Zymo Research Corporation, Irvine, California, USA).
Due to the inavailability of refrigeration, extraction was initiated in the field according to the manufacturers instructions by lysis using a handheld bead-beater (TerraLyser\textsuperscript{TM}; Zymo Research Corporation).

@@

\hypertarget{DNA-supplement}{%
\subsection{DNA extraction, amplification, and sequencing}\label{DNA-supplement}}

After field lysis, DNA was extracted using the Xpedition\textsuperscript{TM} Soil/Fecal Prep kit (see above).
Samples preserved using LifeGuard were first centrifuged at 10~000~g for 1~minute, after which the supernatant was removed and DNA was extracted from the remaining soil using the Soil/Fecal Prep kit as for the other samples.
DNA was quantified fluorometrically using Quant-iT\textsuperscript{TM} PicoGreen\textsuperscript{TM} dsDNA (Thermo Fisher Scientific, Waltham, MA, USA) fluorescent indicator dye on a Infinite F200 plate spectrofluorometer (Tecan Trading AG, Männedorf, Switzerland) according to the manufacturer's protocol.

For the short amplicon, forward primers included sample-specific indexes and adapters for multiplexing (Supplementary File 1).
Amplification was performed by polymerase chain reaction (PCR) in 20µl reactions containing 200~µM dNTP mix, 250~µM indexed gITS7 primer,
150µM ITS4m, 2mM MgCl\textsubscript{2}, 0.1~U \emph{Taq} polymerase (Dream \emph{Taq}, Thermo Fisher Scientific, Waltham, MA, USA) and 3--7~ng purified DNA in Dream \emph{Taq} buffer.
The reaction conditions were 10~min at 95°, followed by 35~cycles of 60~s at 95°, 45~s at 56°, and 50~s at 72°, and finally 3~min at 72°.
Each reaction was conducted in three technical replicates to reduce the effect of PCR stochasticity.

For the long amplicon, both forward and reverse primers included indexes for combinatorical multiplexing (Supplementary File 2).
PCR was performed as for the short amplicons, but with 500~µM of each of the two primers.
Reaction conditions were 10~min at 95°, 30~cycles of 45~s at 95°, 45~s at 59°, and 90~s at 72°, and finally 10~min at 72°.
Each reaction was performed in three technical replicates as for short amplicons.

After pooling of technical replicates, amplicons were purified using SPRI beads \autocite{vesterinen2016} and quantified fluorometrically as above.
An aliquot of 100~ng of DNA from each sample (or the total PCR product if less than 100~ng) was pooled into two libraries each for long and short amplicons.

@@

Raw reads were divided into shorter regions by matching to covariance models (CM), which are similar to stochastic hidden markov models (HMM), but account for both nucleotide sequence and RNA secondary structure \autocite{eddy1994}.
First, the 5.8S rDNA was located in each read by searching for Rfam model RF0002 \autocite{kalvari2018} using \texttt{cmsearch} from Infernal 1.1.2 \autocite{nawrocki2013}, and all bases before the 5.8S were assigned to ITS1.
No attempt was made to remove the approximately 12~bp fragment of the SSU from the 5' end of ITS1 in the long amplicons; it was too short to be reliably detected by a CM or the HMMs employed by ITSx \autocite{bengtsson-palme2013}.
A reference alignment including conserved RNA base pairing between and within the 5.8S and relevant portions of LSU was generated from the fungal 28S RNA seed alignment from RDP release 11.5 \autocite{glockner2017,cole2014} by truncating after the LR5 primer site and using the reference line to annotate the variable regions \emph{sensu} \textcite{michot1984} and \textcite{raue1988}.
A CM was generated from the alignment using \texttt{cmbuild} from Infernal.
The fragment of each read beginning with the 5.8S rDNA was then aligned to the CM using \texttt{cmalign} from Infernal.
The annotation line in the CM alignment for each read was then used to split the reads into alternating more-conserved and less-conserved domains as shown in Figure~\ref{fig:regions},
where LSU1--4 represent the conserved domains of LSU flanking the variable D1--3 domains \autocite{michot1984}.
For short amplicons, only (partial) 5.8S, ITS2, and (partial) LSU1 were extracted.
Code to extract the segments, including annotated seed alignments and CMs, is available in the new R package \texttt{LSUx}, available and documented at \url{https://github.com/brendanf/LSUx}.

@@

\hypertarget{uniform-taxonomic-classification-for-unite-warcup-and-rdp}{%
\subsubsection{Uniform taxonomic classification for Unite, Warcup, and RDP}\label{uniform-taxonomic-classification-for-unite-warcup-and-rdp}}

The taxonomic classification system is based on the Unite database version 8 \autocite{nilsson2019a}.
In particular, the classification for fungi was according to \textcite{tedersoo2018}, and for non-fungal eukaryotes was according to the proposed system of \textcite{tedersoo2017c} as described in \autocite{tedersoodata2017}.
Although the latter system is not formally published, it is consistent with the annotations for non-fungal eukaryotes in the Unite database.
Additionally, it is a system with both purportedly monophyletic taxa and a uniform set of taxon ranks, which make it more appropriate for sequence-based taxonomic assignment algorithms than more accepted classification systems such as that of the International Society of Protistologists \autocite{adl2019}, which utilizes hierarchical nameless ranks.
FASTA-format files for the RDP, Warcup, and Unite reference databases with unified classifications, as well as scripts used to generate them, are available at \url{https://github.com/brendanf/reannotate}.

\hypertarget{rooting-the-fungal-tree}{%
\subsubsection{Rooting the fungal tree}\label{rooting-the-fungal-tree}}

The phylogenetic tree was rooted based on sequences which were confidently assigned to kingdom based on the primary taxonomic assignments.
Assignments based on Warcup were not used at this step because non-Fungi are not included in the dataset, and thus it cannot be used to directly assign kingdoms.
Assignments were considered reliable if at least five of the six remaining primary assignment methods (2 databases × 3 algorithms) resulted in the same kingdom-level assignment, with no conflicting assignments.
Two ASVs belonging to \emph{Tulasnella}, which is known to have an accelerated rate of molecular evolution in the rDNA and often forms long, unstable branches in phylogenies \autocite{moncalvo2006}, were excluded at this step.
This resulted in 516 confident kingdom assignments, with 41 assigned to Alveolata, 453 assigned to Fungi, 1 assigned to Metazoa, and 21 assigned to Viridiplantae.
Inspection of the tree verified that these four kingdoms were reciprocally monophyletic, and the clade containing Alveolata and Viridiplantae was used as an outgroup to root the tree.
The kingdom Fungi was then identified as the minimal clade containing all ASVs which were confidently identified as kingdom Fungi.
ASVs falling outside this clade, including the excluded \emph{Tulasnella} sequences, were not included in downstream fungal community analysis.

\hypertarget{refining-taxonomic-assignments-with-a-phylogenetic-tree-phylotax}{%
\subsubsection{\texorpdfstring{Refining taxonomic assignments with a phylogenetic tree: \texttt{PHYLOTAX}}{Refining taxonomic assignments with a phylogenetic tree: PHYLOTAX}}\label{refining-taxonomic-assignments-with-a-phylogenetic-tree-phylotax}}

@@

\caption[OTU accumulation curves]{OTU accumulation curves. Each curve represents rarefaction of a single sample or pooled pair of samples. Points at the end of each curve represent the actual read depth and observed OTU richness. Note logarithmic axes.}\label{fig:accum-OTU}
\end{figure}

@@

\caption[Comparison of OTU richness between sequencing technologies]{Comparison of OTU richness between sequencing technologies (ref:alpha-compare)}\label{fig:alpha-compare-otu}
\end{figure}

@@

\caption[Heat tree summarizing taxonomic composition of soil community]{Heat tree summarizing taxonomic composition of soil community. Color and size of nodes represent the fractional ASV richness of that taxonomic group.
Color and thickness of branches represent the fractional read abundance of all ASVs belonging to that taxonomic group.
Diverse but relatively rare groups are thus shown as large, dark-colored nodes on relatively thin, light-colored branches (e.g., Dothideomycetes), while common but relatively non-diverse groups are shown as small, light-colored nodes on relatively thick, dark-colored branches (e.g., Russulales).
Groups which are not represented by at least 1\% of reads or 1\% of ASVs in any dataset are collapsed into nodes labeled "*``.
ASVs which could not be further assigned are labelled''?".
ASV richness and read abundance are displayed as the mean across sequencing runs.
Taxonomic assignment is by strict consensus of multiple ITS databases (Unite, Warcup) and assignment algorithms (RDPC, SINTAX, IDTAXA).
The RDP LSU database was not included, to ensure consistent assignment between long and short amplicon libraries.}\label{fig:taxa}
\end{sidewaysfigure}

@@

\caption[Assignment of ECM status to fungal ASVs using FUNGuild database and taxonomic assignments]{Assignment of ECM status to fungal ASVs using FUNGuild database and taxonomic assignments. Reads which were not identified to kingdom, or which were identified as a non-Fungi kingdom, are not included.
``unidentified'' denotes ASVs which were assigned to kingdom Fungi, but could not be identified to family level or could not be assigned ECM status based on the available identification.
Height of each bar represents the fraction of reads represented by that bar.
Results are shown for different sequencing technologies, amplicons (Long, Short), reference databases (Unite, Warcup, RDP; ``All'' denotes algorithms which combine results from multiple databases), and taxonomic assignment algorithms.}\label{fig:ecm}
\end{figure}

@@

\caption[Heat tree showing variation in read abundance and diversity for different amplicons, for ECM taxa only]{Heat tree showing variation in read abundance and diversity for different amplicons, for ECM taxa only. Red-orange edges (nodes) represent increased read (ASV) count for the short amplicon library, while blue-green edges (nodes) represent increased read (ASV) count for the long amplicon library.
Values are log\textsubscript{10} of the ratio of mean values across all sequencing technologies for each amplicon.
Groups which are not represented by at least 1\% of reads or 1\% of ASVs in any dataset are collapsed into nodes labeled "*``.
ASVs which could not be further assigned are labelled''?".
Taxonomic assignment is by strict consensus of multiple ITS databases (Unite, Warcup) and assignment algorithms (RDPC, SINTAX, IDTAXA).
The RDP LSU database was not included, to ensure consistent assignment between long and short amplicon libraries.}\label{fig:ecm-heattree}
\end{figure}

>>>>

% full page landscape figures
\usepackage{rotating}
% \DeclareDelayedFloatFlavour{sidewaysfigure}{figure}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[style=apa,]{biblatex}
\addbibresource{all.bib}
\addbibresource{R.bib}

@@

\textbf{Supplementary file 3}: ML tree of all long metabarcoding ASVs (ITS1--5.8S--ITS2--LSU) assigned to kingdom Fungi.
Thick branches represent rapid bootstrap values \textgreater90\%.
Taxonomic assignment by PHYLOTAX, based on assignments using RDPC, SINTAX, and IDTAXA, with Unite, Warcup, and RDP-LSU reference databases.
Taxon names in blue were not fully resolved with PHYLOTAX, and are given as a list of alternative assignments, along with the number of methods which gave each assignment (max 9).
Taxon names in red were reconstructed as polyphyletic.
In some cases, this may be due to a poorly identified sequence ``splitting'' the group (e.g., the single \emph{Monascus} sequence nested in \emph{Aspergillus}; Aspergillaceae and Trichocomaceae sequences nested in Onygenales), but in other cases, the tree has clearly failed to reconstruct groups which are generally considered to be monophyletic based on multi-gene trees (e.g., Dothidiomycetes, Cantharellales).

@@

\hypertarget{Sampling-supplement}{%
\subsection{Sample preservation and extraction}\label{Sampling-supplement}}

Soil samples described in the main text were placed into separate 2.0~mL microtubes containing 750~mL of field lysis/preservation buffer and lysis beads (Xpedition\textsuperscript{TM} Soil/Fecal DNA miniprep, Zymo Research Corporation, Irvine, California, USA).
Due to the inavailability of refrigeration, extraction was initiated in the field according to the manufacturers instructions by lysis using a handheld bead-beater (TerraLyser\textsuperscript{TM}; Zymo Research Corporation).

@@

After returning to the laboratory, DNA was extracted using the Xpedition\textsuperscript{TM} Soil/Fecal Prep kit (see above).
Samples preserved using LifeGuard were first centrifuged at 10~000~g for 1~minute, after which the supernatant was removed and DNA was extracted from the remaining soil using the Soil/Fecal Prep kit as for the other samples.
DNA was quantified fluorometrically using Quant-iT\textsuperscript{TM} PicoGreen\textsuperscript{TM} dsDNA (Thermo Fisher Scientific, Waltham, MA, USA) fluorescent indicator dye on a Infinite F200 plate spectrofluorometer (Tecan Trading AG, Männedorf, Switzerland) according to the manufacturer's protocol.

@@

Raw reads were divided into shorter regions by matching to covariance models (CM), which are similar to stochastic hidden markov models (HMM), but account for both nucleotide sequence and RNA secondary structure \autocite{eddy1994}.
First, the 5.8S rDNA was located in each read by searching for Rfam model RF0002 \autocite{kalvari2018} using \texttt{cmsearch} from Infernal 1.1.2 \autocite{nawrocki2013}, and all bases before the 5.8S were assigned to ITS1.
No attempt was made to remove the approximately 12~bp fragment of the SSU from the 5' end of ITS1 in the long amplicons; it was too short to be reliably detected by a CM or the HMMs employed by ITSx \autocite{bengtsson-palme2013}.
A reference alignment including conserved RNA base pairing between and within the 5.8S and relevant portions of LSU was generated from the fungal LSU RNA seed alignment from RDP release 11.5 \autocite{glockner2017,cole2014} by truncating after the LR5 primer site and using the reference line to annotate the variable regions \emph{sensu} \textcite{michot1984} and \textcite{raue1988}.
A CM was generated from the alignment using \texttt{cmbuild} from Infernal.
The fragment of each read beginning with the 5.8S rDNA was then aligned to the CM using \texttt{cmalign} from Infernal.
The annotation line in the CM alignment for each read was then used to split the reads into alternating more-conserved and less-conserved domains as shown in Figure~\ref{fig:regions},
where LSU1--4 represent the conserved domains of LSU flanking the variable D1--3 domains \autocite{michot1984}.
For short amplicons, only (partial) 5.8S, ITS2, and (partial) LSU1 were extracted.
Code to extract the segments, including annotated seed alignments and CMs, is available in the new R package \texttt{LSUx}, available and documented at \url{https://github.com/brendanf/LSUx}.

@@

\hypertarget{reannotate}{%
\subsubsection{Uniform taxonomic classification for Unite, Warcup, and RDP-LSU}\label{reannotate}}

The taxonomic classification system is based on the Unite database version 8 \autocite{nilsson2019a}.
In particular, the classification for fungi was according to \textcite{tedersoo2018}, and for non-fungal eukaryotes was according to the proposed system of \textcite{tedersoo2017c} as described in \autocite{tedersoodata2017}.
Although the latter system is not formally published, it is consistent with the annotations for non-fungal eukaryotes in the Unite database.
Additionally, it is a system with both purportedly monophyletic taxa and a uniform set of taxon ranks, which make it more appropriate for sequence-based taxonomic assignment algorithms than more accepted classification systems such as that of the International Society of Protistologists \autocite{adl2019}, which utilizes hierarchical nameless ranks.
FASTA-format files for the RDP-LSU, Warcup, and Unite reference databases with unified classifications, as well as scripts used to generate them, are available at \url{https://github.com/brendanf/reannotate}.

\hypertarget{rooting}{%
\subsubsection{Rooting the fungal tree}\label{rooting}}

The phylogenetic tree was rooted based on sequences which were confidently assigned to kingdom based on the primary taxonomic assignments.
Assignments based on Warcup were not used at this step because non-Fungi are not included in the dataset, and thus it cannot be used to directly assign kingdoms.
Assignments were considered confident if at least five of the six remaining primary assignment methods (2 databases × 3 algorithms) resulted in the same kingdom-level assignment, with no conflicting assignments.
Two ASVs belonging to \emph{Tulasnella}, which is known to have an accelerated rate of molecular evolution in the rDNA and often forms long, unstable branches in phylogenies \autocite{moncalvo2006}, were excluded at this step.
This resulted in 516 confident kingdom assignments, with 41 assigned to Alveolata, 453 assigned to Fungi, 1 assigned to Metazoa, and 21 assigned to Viridiplantae.
Inspection of the tree verified that these four kingdoms were reciprocally monophyletic, and the clade containing Alveolata and Viridiplantae was used as an outgroup to root the tree.
The kingdom Fungi was then identified as the minimal clade containing all ASVs which were confidently identified as kingdom Fungi.
ASVs falling outside this clade, including the excluded \emph{Tulasnella} sequences, were not included in downstream fungal community analysis.

\hypertarget{phylotax}{%
\subsubsection{\texorpdfstring{Refining taxonomic assignments with a phylogenetic tree: \texttt{PHYLOTAX}}{Refining taxonomic assignments with a phylogenetic tree: PHYLOTAX}}\label{phylotax}}

@@

\caption[97\% OTU accumulation curves]{97\% OTU accumulation curves. Each curve represents rarefaction of a single sample or pooled pair of samples.
Points at the end of each curve represent the actual read depth and observed OTU richness.
Panel at left has enlarged scale to show PacBio more clearly.
Vertical dashed line at 100 reads indicates rarefaction level for OTU richness comparisons.}\label{fig:accum-OTU}
\end{figure}

@@

\caption[Comparison of OTU richness between sequencing technologies]{Comparison of OTU richness between sequencing technologies Each point represents the richness of one or two pooled samples, as determined by two different sequencing strategies. All values represent the average of 100 replicate rarefactions with a sample depth of 100 reads. Because samples in the same well on different plates could not be demultiplexed in the Ion Torent dataset, these samples were also bioinformatically pooled in the other datasets prior to rarefaction. Blue lines are total least squares fit with 95\% confidence interval, with the given slope (and 95\% confidence interval) and R\textsuperscript{2} value. Dashed diagonal line indicates ideal slope of 1.}\label{fig:alpha-compare-otu}
\end{figure}

@@

\caption[Heat tree summarizing taxonomic composition of soil community]{Heat tree summarizing taxonomic composition of soil community. Color and size of nodes represent the fractional ASV richness of that taxonomic group.
Color and thickness of branches represent the fractional read abundance of all ASVs belonging to that taxonomic group.
Diverse but relatively rare groups are thus shown as large, dark-colored nodes on relatively thin, light-colored branches (e.g., Dothideomycetes), while common but relatively non-diverse groups are shown as small, light-colored nodes on relatively thick, dark-colored branches (e.g., Russulales).
Groups which are not represented by at least 1\% of reads or 1\% of ASVs in any dataset are collapsed into nodes labeled "*``.
ASVs which could not be further assigned are labelled''?".
ASV richness and read abundance are displayed as the mean across sequencing runs.
Taxonomic assignment is by strict consensus of multiple ITS databases (Unite, Warcup) and assignment algorithms (RDPC, SINTAX, IDTAXA).
The RDP-LSU database was not included, to ensure consistent assignment between long and short amplicon libraries.}\label{fig:taxa}
\end{sidewaysfigure}

@@

\caption[Assignment of ECM status to fungal ASVs using FUNGuild database and taxonomic assignments]{Assignment of ECM status to fungal ASVs using FUNGuild database and taxonomic assignments. Reads which were not identified to kingdom, or which were identified as a non-Fungi kingdom, are not included.
``unidentified'' denotes ASVs which were assigned to kingdom Fungi, but could not be identified to family level or could not be assigned ECM status based on the available identification.
Height of each bar represents the fraction of reads represented by that bar.
Results are shown for different sequencing technologies, amplicons (Long, Short), reference databases (Unite, Warcup, RDP-LSU; ``All'' denotes algorithms which combine results from multiple databases), and taxonomic assignment algorithms.}\label{fig:ecm}
\end{figure}

@@

\caption[Heat tree showing variation in read abundance and diversity for different amplicons, for ECM taxa only]{Heat tree showing variation in read abundance and diversity for different amplicons, for ECM taxa only. Red-orange edges (nodes) represent increased read (ASV) count for the short amplicon library, while blue-green edges (nodes) represent increased read (ASV) count for the long amplicon library.
Values are log\textsubscript{10} of the ratio of mean values across all sequencing technologies for each amplicon.
Groups which are not represented by at least 1\% of reads or 1\% of ASVs in any dataset are collapsed into nodes labeled "*``.
ASVs which could not be further assigned are labelled''?".
Taxonomic assignment is by strict consensus of multiple ITS databases (Unite, Warcup) and assignment algorithms (RDPC, SINTAX, IDTAXA).
The RDP-LSU database was not included, to ensure consistent assignment between long and short amplicon libraries.}\label{fig:ecm-heattree}
\end{figure}
